---
title: "KODAMA: Knowledge Discovery by Accuracy Maximization"
author: "Stefano Cacciatore"
package: KODAMA
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{KODAMA: Knowledge Discovery by Accuracy Maximization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

`KODAMA` is the R interface to the standalone `kodama-cpp` library. KODAMA
constructs a reusable nearest-neighbor graph, evolves independent candidate
label vectors by maximizing cross-validated prediction accuracy, and uses the
ensemble of label solutions to correct graph dissimilarities for downstream
embedding. The supported classifiers are KNN and SIMPLS followed by LDA in
latent-component space.

The numerical core uses float32 data and has CPU, NVIDIA CUDA, and Apple Metal
backends. The R interface deliberately remains focused on the KODAMA workflow;
general dimensionality-reduction and PLS modeling APIs are maintained in
`fastEmbedR` and `fastPLS`.

# Basic workflow

```{r data}
library(KODAMA)
data(MetRef)

x <- MetRef$data
x <- x[, colSums(x) != 0, drop = FALSE]
x <- normalization(x)$newXtrain
x <- scaling(x)$newXtrain
```

Build the expensive graph once and reuse it for either classifier:

```{r workflow, eval=FALSE}
g <- KODAMA.graph(x, k = 100, backend = "cpu", n.cores = 4)

fit.knn <- KODAMA.matrix(
  data = x, graph = g, classifier = "knn",
  M = 100, Tcycle = 100, knn.k = 30,
  backend = "cpu", n.cores = 4
)

fit.pls <- KODAMA.matrix(
  data = x, graph = g, classifier = "pls_lda",
  M = 100, Tcycle = 100, ncomp = 50,
  backend = "cpu", n.cores = 4
)

embedding <- KODAMA.visualization(
  fit.pls, method = "UMAP", raw.data = x,
  graph.mode = "fuzzy", backend = "cpu"
)
```

`KODAMA.matrix()` also accepts raw data without an explicit graph. In that
case it calls `KODAMA.graph()` internally. A precomputed KNN graph can be used
when the original feature matrix is unavailable; PLS-LDA then operates on
self-tuning graph-Laplacian features.

# Bioconductor objects

`RunKODAMAgraph()`, `RunKODAMAmatrix()`, and `RunKODAMAvisualization()` provide
methods for `SingleCellExperiment`, `SpatialExperiment`, Seurat, and Giotto
objects. Optional packages are checked at run time and remain in `Suggests`, so
the base package can be installed without a complete single-cell ecosystem.

# Migration from KODAMA 2.4

This package is a new implementation, not a byte-for-byte continuation of the
legacy R code. The previous implementation is archived as `KODAMAlegacy`.

## Innovations

* The mathematical core is a standalone C++17 library that can be wrapped by R
  and Python without embedding either language runtime.
* CPU, CUDA, and Metal backends share the same KODAMA evolution contract and
  use float32 analysis buffers.
* `KODAMA.graph()` separates graph construction from accuracy maximization, so
  KNN and PLS-LDA analyses can reuse the same graph and PCA initialization.
* `KODAMA.matrix()` accepts raw matrices, reusable graph handles, or explicit
  KNN indices and distances.
* KNN and SIMPLS plus latent-space LDA are explicit classifiers. Independent
  `M` runs retain cycle diagnostics, timing, and backend provenance.
* Landmark strata are constructed once as a shared coarse atlas, preserving
  coverage while avoiding repeated full-data partitioning.
* Native UMAP and openTSNE consume the corrected KODAMA graph and use raw-data
  PCA initialization by default when available.

## Renamed or deprecated interfaces

| Legacy interface | New location or replacement |
|---|---|
| `KODAMA.matrix()` | Retained with graph reuse, explicit classifier, backend, and diagnostics arguments. |
| `KODAMA.visualization()` | Retained; configuration is supplied as named arguments rather than `config.*.default` lists. |
| `core_cpp()` | Replaced by `CoreKNN()` and `CorePLSLDA()`. |
| `pca()` | Use `fastEmbedR::pca()` for the general R workflow. `KODAMA.pca()` remains available for backend-native initialization and diagnostics. |
| t-SNE/UMAP configuration objects | Use `fastEmbedR` for general embedding or named arguments to `KODAMA.visualization()`. |
| legacy PLS-DA helpers | Use `fastPLS::pls()` for general PLS modeling; KODAMA supports KNN and PLS-LDA classifiers. |
| `floyd()`, `transformy()`, `kabsch()`, `mcplot()` | Deprecated legacy utilities. Diagnostics are available through `KODAMA.diagnostics()` and `KODAMA.timing()`. |
| `ancestry = TRUE` | Replaced by the explicit `spatial.mode = "population"`. |

The synthetic manifold generators, `normalization()`, and `scaling()` remain
available for compatibility. The historical `MetRef`, `USA`, and `lymphoma`
datasets are included so published examples remain reproducible.
