---
title: "KODAMA: Knowledge Discovery by Accuracy Maximization"
author: "Stefano Cacciatore"
package: KODAMA
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{KODAMA: Knowledge Discovery by Accuracy Maximization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(KODAMA)
```

# Overview

`KODAMA` is the R interface to the standalone `kodama-cpp` library. KODAMA
constructs a reusable nearest-neighbor graph, evolves independent candidate
label vectors by maximizing cross-validated prediction accuracy, and uses the
ensemble of label solutions to correct graph dissimilarities for downstream
embedding. The supported classifiers are KNN and SIMPLS followed by LDA in
latent-component space.

The numerical core uses float32 data and has CPU, NVIDIA CUDA, and Apple Metal
backends. The R interface deliberately remains focused on the KODAMA workflow;
general dimensionality-reduction and PLS modeling APIs are maintained in
`fastEmbedR` and `fastPLS`.

# Select default execution resources

Backend-capable functions use CPU unless a backend is selected for the R
session. Pipeline functions also use a session-wide core count when supplied.
Explicit function arguments have the highest priority.

```{r backend-configuration, eval=TRUE}
# Session-wide defaults
old.backend <- KODAMA_backend("cpu")
old.cores <- getOption("n.cores")
options(n.cores = 4L)

# The equivalent direct option is options(backend = "cpu")
KODAMA_backend()

# Explicit arguments override the session default. For example:
# KODAMA.graph(x, backend = "cuda")

options(backend = old.backend)
options(n.cores = old.cores)
```

For an accelerator session, set one of the following before the analysis:

```{r accelerator-configuration, eval=FALSE}
KODAMA_backend("cuda")
KODAMA_backend("metal")
options(n.cores = 4L)

# Uses the session default
graph <- KODAMA.graph(x)
```

For non-interactive jobs, `BACKEND=cpu|cuda|metal` supplies the default
when the R option is unset. The precedence is explicit argument, R option,
environment variable, then CPU. Selecting an unavailable accelerator produces
an error; KODAMA does not silently change the requested backend.

# Basic workflow

```{r data}
data(MetRef)

x <- MetRef$data
x <- x[, colSums(x) != 0, drop = FALSE]
x <- normalization(x)$newXtrain
x <- scaling(x)$newXtrain
```

Build the expensive graph once and reuse it for either classifier:

```{r workflow, eval=FALSE}
g <- KODAMA.graph(x, k = 100, backend = "cpu", n.cores = 4)

fit.knn <- KODAMA.matrix(
  data = x, graph = g, classifier = "knn",
  M = 100, Tcycle = 100, knn.k = 30,
  backend = "cpu", n.cores = 4
)

fit.pls <- KODAMA.matrix(
  data = x, graph = g, classifier = "pls_lda",
  M = 100, Tcycle = 100, ncomp = 50,
  backend = "cpu", n.cores = 4
)

embedding <- KODAMA.visualization(
  fit.pls, method = "UMAP", raw.data = x,
  graph.mode = "fuzzy", backend = "cpu"
)
```

`KODAMA.matrix()` also accepts raw data without an explicit graph. In that
case it calls `KODAMA.graph()` internally. A precomputed KNN graph can be used
when the original feature matrix is unavailable; PLS-LDA then operates on
self-tuning graph-Laplacian features.

# Bioconductor objects

`RunKODAMAgraph()`, `RunKODAMAmatrix()`, and `RunKODAMAvisualization()` provide
methods for `SingleCellExperiment`, `SpatialExperiment`, Seurat, and Giotto
objects. Optional packages are checked at run time and remain in `Suggests`, so
the base package can be installed without a complete single-cell ecosystem.

The following compact workflow keeps the analysis and its reusable graph inside
a `SpatialExperiment`. It replaces the legacy extraction of separate data,
coordinate, sample, and label objects. `sample_id` and `spatialCoords()` are
read automatically by the object methods.

```{r spatialexperiment-workflow, eval=FALSE}
spe <- SpatialFeatureSelection(spe)
subject <- "Br5595"
spe_sub <- spe[, spe$subject == subject]
spe_sub <- RunFastPCA(
  spe_sub, nfeatures = gene_number, ncomp = 50,
  center = TRUE, scale = TRUE,
  backend = backend, n.cores = n.cores, seed = seed
)

spe_sub <- RunKODAMAgraph(
  spe_sub, dims = 50, backend = backend, n.cores = n.cores
)
spe_sub <- RunKODAMAmatrix(
  spe_sub, dims = 50, classifier = "pls_lda", M = 100, Tcycle = 100,
  landmarks = 100000, splitting = splitting,
  spatial.resolution = spatial.resolution,
  backend = backend, n.cores = n.cores, seed = seed
)
spe_sub <- RunKODAMAvisualization(
  spe_sub, dims = 50, method = "UMAP",
  backend = backend, n.cores = n.cores
)

labels <- factor(spe_sub$layer_guess_reordered)
plot(
  SingleCellExperiment::reducedDim(spe_sub, "KODAMA"),
  pch = 20, col = cols_cluster[labels]
)
```

# Migration from KODAMA 2.4

This package is a new implementation, not a byte-for-byte continuation of the
legacy R code. The previous implementation is archived as `KODAMAlegacy`.

## Innovations

* The mathematical core is a standalone C++17 library that can be wrapped by R
  and Python without embedding either language runtime.
* CPU, CUDA, and Metal backends share the same KODAMA evolution contract and
  use float32 analysis buffers.
* `KODAMA.graph()` separates graph construction from accuracy maximization, so
  KNN and PLS-LDA analyses can reuse the same graph and PCA initialization.
* `KODAMA.matrix()` accepts raw matrices, reusable graph handles, or explicit
  KNN indices and distances.
* KNN and SIMPLS plus latent-space LDA are explicit classifiers. Independent
  `M` runs retain cycle diagnostics, timing, and backend provenance.
* Landmark strata are constructed once as a shared coarse atlas, preserving
  coverage while avoiding repeated full-data partitioning.
* Native UMAP and openTSNE consume the corrected KODAMA graph and use raw-data
  PCA initialization by default when available.

## Renamed or deprecated interfaces

| Legacy interface | New location or replacement |
|---|---|
| `KODAMA.matrix()` | Retained with graph reuse, explicit classifier, backend, and diagnostics arguments. |
| `KODAMA.visualization()` | Retained; configuration is supplied as named arguments rather than `config.*.default` lists. |
| `core_cpp()` | Replaced by `CoreKNN()` and `CorePLSLDA()`. |
| `pca()` | Use `RunFastPCA()` for backend-native float32 PCA on matrices or supported containers. |
| `config.tsne.default` and `config.umap.default` | Use `fastEmbedR` for general embedding or named arguments to `KODAMA.visualization()`. |
| `MDS.defaults` and the legacy MDS visualization path | Deprecated. Use the graph-based UMAP or openTSNE methods in `KODAMA.visualization()`, or a dedicated MDS implementation when metric MDS is specifically required. |
| legacy PLS-DA helpers | Use `fastPLS::pls()` for general PLS modeling; KODAMA supports KNN and PLS-LDA classifiers. |
| `floyd()`, `transformy()`, `kabsch()`, `mcplot()` | Deprecated legacy utilities. Diagnostics are available through `KODAMA.diagnostics()` and `KODAMA.timing()`. |
| `ancestry = TRUE` | Replaced by the explicit `spatial.mode = "population"`. |

The synthetic manifold generators, `normalization()`, and `scaling()` remain
available for compatibility. The historical `MetRef`, `USA`, and `lymphoma`
datasets are included so published examples remain reproducible.

# Session information

```{r session-info}
sessionInfo()
```
